Retirement Consumption: Evidence from UK Pension Reform

### 2023-07-27 Texfiles/Draft1.tex
almost final commit

        Word count: [Insert]
        \vfill
        I confirm that this is entirely my own work and has not previously been submitted for assessment, and I have read and understood the University’s and Faculty’s definition of Plagiarism
    \end{center}
\end{titlepage}
\newpage
\maketitle
\begin{abstract}
    I identify the consumption responses of retirees to the UK's `pensions
    freedom act', a landmark law change that meant defined contribution pension
    holders were no longer forced to annuitise their funds and could instead
    access their pensions in a variety of ways. I use a discontinuity in year of
    retirement and a difference-in-difference approach, harnessing those who
    were unaffected by the policy reform as a natural control group. I then
    solve three sets of lifecycle models, tracking assets from retirement until an
    agent reaches age 110. I simulate their consumption with and without
    annuities, mimicking the policy reform, and match their consumption
    responses to what was seen in the empirical models. I find that a lifecycle
    model with a bequest motive best fits the identified consumption response —
    a result that could inform policymakers in designing optimal pension
    policies for an ageing population.
\end{abstract}
\newpage
\tableofcontents
\newpage

@@

I then test two competing hypotheses for the annuity puzzle: bequests and
pessimistic life expectancy. The annuity puzzle was first identified by
\cite{yaari_65} and is a classed as such since standard economic theory cannot
explain the lack of annuitisation that is observed in the real world. Two
possible explanations are bequests and life expectancy: individuals want to
leave an estate for their heirs to inherit and they cannot do this if they
annuitise their wealth; or individuals may believe that they will not live as
long as annuity providers think they will, and therefore annuities that are on
the market do not appear to be good value. Depending on the reason for the lack
of annuitisation in the UK, the consumption response of retirees to the pension
reform will differ. If individuals do not annuitise because of pessimistic life
expectancy, their consumption should increase after the reform. If, on the other
hand, individuals do not annuitise because of a bequest motive, consumption
should not change as much as a result of the reform.

I will solve lifecycle models for both of these cases and simulate consumption
decisions with and without forced annuitisation. I will then use a variety of
empirical models to measure the consumption change in early retirement that
resulted from the policy reform. The sign and magnitude of this change will be
indicative of the mechanism causing the annuitisation problem.

The importance of retirement policy to individuals in the UK is growing. The
number of individuals of pensionable age is expected to increase from 11.9
million in 2020 to 15.2 million in 2045 according to the latest projections from
the Office for National Statistics (ONS), and for every 1,000 people of working
age there will be 341 of pensionable age in 2045 compared to 280 in 2020
\citep{ons_population_predictions_2020}. The increase in absolute and relative
numbers of elderly people makes retirement policy more consequential. Moreover,
private DC pensions are becoming increasingly common and are predicted to grow
as current cohorts age \citep{cribb_karjalainen_ifs_2023}. Therefore, policies
regarding how private pensions can be accessed will have a larger impact on
overall welfare for retirees.

Moreover, understanding how retirees spend their money over retirement is an
important policy question of its own. If retirees spend too much in early
retirement, the state may need to provide for them towards the end of their
lives. This has implications for government budgets, especially since population
aging means more individuals will require expensive end of life care. On the
other hand, if retirees over-save and do not spend, the economy may be
dynamically inefficient -- when the capital stock is too high -- because
individuals save too much for retirement and for bequests. In this case
consumption per capita could be increased if savings were decreased.

\subsection{Literature review}
My paper draws on three main strands of literature: that of the annuity puzzle,
exploring reasons why retirees choose not to annuitise; the retirement
saving puzzle, which seeks to explain why retirees drawdown assets slowly; and
lifecycle models, whereby the above problems are explained using a tractable
model of human decision making.

\cite{yaari_65} showed that under standard assumptions with risk averse agents
we would expect individuals to annuitise all of their wealth at retirement to
insure against the risk of long life. More recently,
\cite{davidoff_brown_diamond_aer_2005} show that complete annuitisation is
optimal under a broader range of assumptions. Since then, there has been much
discussion of possible reasons for why people choose not to annuitise.
\cite{finkelstein_porteba_2002, finkelstein_porteba_2004} find evidence of
adverse selection into the UK annuity market and \cite{mitchell_et_al_1999} find
evidence of the same in the US. This means longer-lived individuals select into
buying annuities which drives up annuity prices because providers need to pay
individuals for longer than they originally predicted they would need to. This
makes the `money's worth' of an annuity lower for the general population as
opposed to the population of annuitants, putting off individuals from buying an
annuity. However, \cite{mitchell_et_al_1999} also find that theory would still
predict annuitisation because the money's worth of annuities is not that far
from `actuarially fair' (meaning the present discounted value of annuities
equals the price that an individual pays for it).

\cite{friedman_warshawsky_qje_1990} show that annuitisation decisions can be
fully explained by a mixture of bequest motives and `actuarially unfair'
annuities. They solve an augmented lifecycle model and calculate annuity demand
for a range of differences between the rate of return assets would get in the stock
market and the rate of return annuities give an individual. For plausible values,
they find that individuals would optimally not annuitise much wealth. Similarly
to \cite{finkelstein_porteba_2004}, \cite{friedman_warshawsky_chicago_1988} show
that there is a significant difference between the life expectancy of annuitants
and the general population in the American annuity market, which therefore
implies a degree of adverse selection. But, they find this cannot fully explain
the annuitisation puzzle, and only when bequest motives are added to the model
can annuitisation rates be rationalised.

More recently, \cite{lockwood_red_2012} builds on this and shows that a
realistic bequest motive in lifecycle simulations achieves realistic
annuitisation rates. He solves a simple lifecycle model with bequest motives
taken from several recent papers in the retirement savings literature. The
bequest motives he picks match other important aspects of the empirical
distribution of bequests, such as how much individuals bequeath and the financial
characteristics of these individuals. Also, \cite{odea_sturrock_rest_2023} find
that subjective life expectancies can explain a large part of the annuitisation
puzzle, especially if agents are less risk averse.

Apart from bequests and annuity prices, a range of other reasons for the annuity
puzzle have been explored. Precautionary savings for unexpected end-of-life
medical expenses might cause individuals to prefer holding assets rather than
an annuity. Moreover, \cite{vidalmelia_lejarragagarcia_munich_2004} suggest
that couples might share the risk of long life by pooling their resources, and
therefore couples may be less likely to annuitise their wealth.

Given that uncertain end-of-life medical expenses are not such an issue in the
UK as opposed to the US, and that the focus of recent literature has been on the
bequest model and uncertain life expectancy, I choose also to focus on bequest
and objective life expectancy in the UK context. To my knowledge, there has been
no attempt to analyse the consumption response to 2014 reform, nor evaluate what
common lifecycle models predict for the policy change.

\section{Data and Policy reform}

@@

\textbf{I could go into more detail here? Maybe I should. Add equation etc}

@@

Successive governments in the UK have attempted to reform both the public and
private pension system. Prior to 1987, participation in private schemes was
limited to employees of firms that had offered them, and there were few
alternatives to the public state pension or DB scheme that a public sector
employer would offer. From 1989, individuals in the UK were allowed to open tax
advantaged self-invested personal pensions alongside any pension their employer
offered. The 2004 Finance Act rationalised taxation rates on DB and DC pensions.
The rules for DC pension pots were such that an individual had to buy an annuity
after optionally taking a maximum of 25\% of the pot as a tax-free lump sum
withdrawal. DC pension pots were accessible from age 55, and most required that
they be accessed before the age of 75.

@@

\textbf{For conclusion: future research could explore the impact of this second reform on the
    consumption and saving decisions of retirees.}

The impact of the reforms on annuity demand has been documented by
\cite{cannon_et_al_nier_2016}. Using data from the Association of British
Insurers, they show that annuity demand dropped by 75\% from its maximum in
2011 to 2016. Figure \ref{fig:annovertime} shows
purchases of annuities over time and demonstrates the sharp decrease in purchases
that ocurred from 2014 to 2015. There was an increase in the number of pensions
that were being accessed using an income drawdown product, but these only partly
account for the drop in annuitisation as is shown by the graph.
\begin{figure}[h]

    \centering
    \includegraphics[width=0.9\columnwidth]{figures/annuity_overtime.pdf}
    \label{fig:annovertime}
    \caption{New pension pots accessed}
\end{figure}

@@

    \
\end{figure}
Figure \ref{fig:ann2122} shows the distribution of methods used to access pension pots
at retirement in 2021-22. In 2021, 196,736 pots were fully withdrawn at
retirement, accounting for over 50\% of pots. Prior to the policy reform this
was not the case -- most DC pensions were accessed through
annuities.

Also of specific interest to the annuity market was the European Union's `Equal
Treatment in Goods and Services Directive' of 2004. This prohibited
discrimination in the provision and cost of goods and services based on sex. Up
until 2011, there was a clause that stated insurers were allowed to charge
different premiums if this was based on evidence that sex is correlated with
different amounts of risk. However, in March 2011, the European Court of Justice
ruled that insurers were not allowed to charge different amounts and gave them
until December 2012 to implement the ruling. This change meant that annuity
products could no longer be priced differently for men and women in the UK.
However, given that this change was implemented two years before the pension
freedoms act it will not impact my results.

@@

\subsection{Covariate Balance}

Table \ref{tab:sum_stats}\footnote{There are hyperlinks on the table number so,
    if your pdf viewer supports this, you can click to go to the table or
    figure.} shows various summary statistics for each variable of interest from
ELSA and the ONS.

As required, retirement year is between 2011 and 2013 for the control group and
2015 and 2017 for the treatment group. However, as discussed above their ELSA
interview date are after or the same as the retirement year because we are
tracking consumption in retirement.

The second retirement group, those who retired post-reform, is small with just
728 non missing observations for gender as opposed to 941 individuals in the
control group. The control group are more female, retired at a slightly younger
age and expected to retire slightly earlier. This could be because this period
was affected by the increase in the state pension age for women from age 60 in
2010 to age 65 in 2018. Since this change happened gradually and ocurred over
the whole period I do not expect it to influence the results. The DifferenceAge
row tracks the difference between expected and actual retirement age.

The treatment group has higher financial wealth with a median of £67,000 at the
time of interview as opposed to £55,000 in the control group. Likewise, the
second group are more likely to have held a DC pension at some point and have
more money in them. Both groups are equally likely to have a DB pension with
roughly 45\% of individuals across both samples having a DB pension. In general,
DB pensions are more prevalent than DC pensions in the data -- this reflects the
trend that is observed in the general UK population. Home ownership is roughly
equal across groups although the second group has slightly higher housing
wealth.

@@

Table \ref{tab:cov_balance} shows these results. As expected, retirement year
and interview year are greater for the treated group. The difference between
expected and real retirement age is 0 which shows that individuals are not
delaying retirement, beyond their original expectation,  in order to be part of
the treatment group although retirement age is higher post reform. DC pension
value is higher, as is financial wealth but these are both imprecisely
estimated. Because of the differences in these variables I add them to the
regression model.

One threat to validity is manipulation of the running variable, retirement year.
This is probably quite a threat in this setting as individuals could delay
retirement and/or the purchase of an annuity until after the reform. As observed
above in Tables \ref{tab:sum_stats} and \ref{tab:cov_balance}, it does not
appear as though individuals are delaying retirement. However, it is possible
that individuals do not buy an annuity at the time of retirement since the law
prior to 2014 only required that they access the pot by age 75. They may decide
to keep their DC pension untouched and live off other income and assets before
accessing it later. Since we do not have data on the exact purchase date of
annuities, we cannot track whether annuity purchases were delayed because of
this. However, there is relatively stable annuity demand before the policy
reform. If individuals delayed annuity purchases in expectation of the reform,
so that they would not have to buy an annuity at all, we would observe declines
in purchases before the reform was announced. Likewise, if individuals bought an
annuity early in expectation that the end of the compulsory market would cause
prices to go up, we would observe a spike in annuity purchases prior to 2014.

@@

In this section, I outline the key empirical models I run with first the real
data from ELSA, and then the simulated data from the lifecycle models. I then
see which lifecycle model better fits the consumption response that occurred as
a result of the pension reform.

I use a fuzzy, since some individuals will probably still annuitise a little,
regression discontinuity design for which I compare the consumption of
individuals who retired after the policy reform to that of people who retired
just before the reform. The key assumption implicit in regression
discontinuities is that nothing else changes at the time of the jump apart from
the policy of interest, and that the policy occurs without individuals
predicting it and altering their behaviour. As I have argued above, the
demographic information in the data suggests that individuals did not delay
retirement and that there was no delay in annuitisation as shown by the quick
and sudden decline in annuity purchases. Moreover, anecdotal evidence from media
and business sources at the time of the announcement show that there was
surprise the government put forward the reform, with Money Marketing describing
the change as a ``bombshell" \citep{money_marketing_announcement}.

Retirement year is the running variable and individuals are treated if
retirement year is greater than 2014 and less than or equal to 2017. I use
consumption data of individuals up to 2 years into retirement so that the sample
size is larger. So, if someone retired in 2015 and had consumption data in 2015
and 2017, I include both values. An individual is considered not treated if they
retire in or before 2013.

@@

Table \ref{tab:DcOnlyRes} shows the results of this specification on the various
consumption indicators. Column 1 has total monthly consumption on the left hand
side of the regression equation. We can see that being in the treatment group is
associated with £36 more a week in spending overall across all categories of
expenditure. This is not statistically significant using robust standard errors.
Being further into retirement at the time of interview is associated with higher
total consumption whilst those who retired later have lower consumption. This is
surprising since lifecycle models predict consumption decreases over retirement
and this has also been documented empirically \citep{hurd_rohwedder_nber_2003}.

Columns 2 through 4 use food consumption as the outcome variable. Interestingly,
food consumption inside the house does not increase as a result of the policy
reform but food consumed away from the house does increase. A shock to income,
caused by no longer needing to annuitise, may cause spending on luxury goods to
increase and for spending on their substitutes, like food in the house relative
to food away from home, to decrease. This income effect may be driving the
results we observe in Table \ref{tab:DcOnlyRes} since food consumption outside the
home increases as does expenditure on clothing -- seen in column 5.

@@

The key identifying assumption for DD is the parallel trends assumption. In this
context, it means that consumption differences between the DB, DC and no pension
group across cohorts would have been the same were it not for the policy change.
For DD to not be valid we would need a reason to expect that the individuals
with a DB and no pension who are retiring post 2014 are impacted by some factor
that did not equally impact the DC group and that would impact the consumption
of all groups differently. Given that there were no other reforms to pension
schemes at this time that would have only impacted one of these groups in
particular, I claim that the parallel trends assumption holds holds.

Table \ref{tab:ElsaAllData} shows these results. The post-reform variable now
tells us that total and food consumption increased but clothing consumption
decreased. The interactions at the bottom of the table show us what happened to
individuals with a given pension type relative to the reference cohort which has
no pension. Those with only a DC pension had an increase in total consumption
which was higher than those with a DB pension and higher than those with no
pension. Consumption on food decreased slightly for the DC group relative to no
pension, but the decrease was less than that for the DB group. Expenditure on
clothing also increased for the DC group compared to no pension and DB pensions.
The increase in sample size also brings decreases in the standard errors of
these variables -- the interactions are now all significant a the 5\% level as
is the main coefficient on retiring post-reform.

However, it is worth noting that there is a significant change between the DB
and the no pension group. A natural test of the parallel trends assumption is to
see whether two different, untreated, units have different outcomes
post-treatment. If they are different post treatment then there could be
something else driving the differences between the treatment group and the
non-treated groups.

We can also test whether treatment intensity is correlated with larger increases
in spending. Those who have more money in a DC pension would be impacted by the
policy to a greater extent than those with smaller pensions, since they would be
forced to annuitise a larger amount. So, I interact the policy with financial
wealth and DC wealth. As in Table \ref{tab:DcOnlyRes}, this regression only uses
those individuals who have a DC pension.

@@

\begin{landscape}
    \linespread{1}

    \input{tables/elsa_results_only_dc.tex}
\end{landscape}

\begin{landscape}
    \linespread{1}

    \input{tables/elsa_results_all_with_interact.tex}
\end{landscape}

\begin{landscape}
    \linespread{1}

    \input{tables/elsa_results_dc_pot_interact.tex}
\end{landscape}

\begin{landscape}
    \linespread{1}
    \input{tables/elsa_results_fin_wealth_interact.tex}
\end{landscape}

Table \ref{tab:DcOnlyInteract} shows the results. The interaction with DC
pension size is positive, meaning that a larger DC pension pot associated with a
greater increase in total monthly consumption. But, although the reform is again
associated with an increase in consumption, especially for food out of the house
and clothing, this effect is decreasing with the size of the pension pot for
these consumption goods.

The DC pension pot size variable comes from work done by the IFS. The data after
wave 5 has not been released yet, so for individuals who retire after 2011 I
have no DC wealth variable. For Table \ref{tab:DcOnlyInteract}, pension wealth
from the last available year for an individual was given a real return of 3\%
and compounded until year of interview. Given that this is an imperfect measure,
I also run the regression with financial wealth interacted with the treatment
variable.

Table \ref{tab:DcOnlyFinWealthInteract} contains these results and shows a
similar pattern to Table \ref{tab:DcOnlyInteract}. The reform is associated with
an increase in total monthly consumption and more financial wealth increases.

\subsection{Robustness Checks}

I carry out two checks for robustness. Firstly, I exclude those individuals who
retired more than one year away from their expected retirement year -- I define
an individuals expected retirement year as their expected retirement age in
their first interview with ELSA, added to the year that the interview was
carried out in. This checks whether the effects shown above are being driven by
individuals either retiring early in order to buy cheaper annuities or retiring
later and avoiding annuitisation altogether.

I also check whether the EU reform to annuity pricing is changing the results in
some way. Since the reform only came into affect in December 2012 there are some
retirees in our sample who may have been able to purchase an annuity before
annuity pricing was made the same for men and women. I drop anyone with a
retirement year in 2012 or 2011 and repeat the analysis from Table
\ref{tab:DcOnlyRes}.

\begin{landscape}
    \linespread{1}
    \input{tables/rob_elsa_results_only_dc_only_exp.tex}
\end{landscape}

@@

Tables \ref{tab:DcOnlyExpOnlyRes} and \ref{tab:DcOnlyNot2012} show these
results. In Table \ref{tab:DcOnlyExpOnlyRes}, we can see the sign is reversed on
PostReform in all consumption categories apart from clothing, however the sample
size is a third of what it was prior to excluding individuals so there is more
uncertainty in these estimates. In Table \ref{tab:DcOnlyNot2012}, the sample
size is again much smaller and the sign has flipped on all consumption
categories apart from clothing and food consumed in the house. These regressions
show that there are potentially identification issues with the regression
discontinuity and DD strategies that I have presented thus far. However, since
the sample sizes for both the robustness tests and main results are small it is
difficult to asses what the impact of the reform was.

>>>>

        Word count: 8335\footnote{This is an estimate that my LaTeX editor gave me.} + 1750 (figure and table word equivalent)
        \vfill
        I confirm that this is entirely my own work and has not previously been submitted for assessment, and I have read and understood the University’s and Faculty’s definition of Plagiarism
    \end{center}
\end{titlepage}
\newpage
\maketitle
\begin{abstract}
    I identify the consumption responses of retirees to the UK's `pensions
    freedom act', a landmark law change that meant defined contribution pension
    holders were no longer forced to annuitise their funds and could instead
    access their pensions in a variety of ways. I use a discontinuity in year of
    retirement and a difference-in-difference approach, harnessing those who
    were unaffected by the policy reform as a natural control group. I then
    solve three sets of lifecycle models, tracking assets from retirement until an
    agent reaches age 110. I simulate their consumption with and without
    annuities, mimicking the policy reform, and match their consumption
    responses to what was seen in the empirical models. I find that a lifecycle
    model with a bequest motive best fits the identified consumption response —
    a result that could inform policymakers in designing optimal pension
    policies for an ageing population.
\end{abstract}
\newpage
\tableofcontents
\newpage

@@

I then test two competing hypotheses for the annuity puzzle: bequests and
pessimistic life expectancy. A bequest motive means that individuals want to
leave an estate for their heirs to inherit -- which they cannot do if they
annuitise their wealth. On the other hand, individuals may believe that they
will not live as long as annuity providers think they will, and therefore
annuities that are on the market do not appear to be good value. Depending on
the reason for the lack of annuitisation in the UK, the consumption response of
retirees to the pension reform will differ. If individuals do not annuitise
because of pessimistic life expectancy, their consumption should increase after
the reform. If, on the other hand, individuals do not annuitise because of a
bequest motive, consumption should not change as much as a result of the reform.

I will solve lifecycle models for both of these cases and simulate consumption
decisions with and without annuitisation. I will then apply the empirical models
to this simulated data and measure the consumption change in early retirement
that results from annuitisation. The sign and magnitude of these coefficients
relative to the coefficients from the real data will be indicative of the
mechanism causing the annuitisation puzzle.

The importance of retirement policy to the UK is growing. The
number of individuals of pensionable age is expected to increase from 11.9
million in 2020 to 15.2 million in 2045 according to the latest projections from
the Office for National Statistics (ONS), and for every 1,000 people of working
age there will be 341 of pensionable age in 2045 compared to 280 in 2020
\citep{ons_population_predictions_2020}. The increase in absolute and relative
numbers of elderly people makes retirement policy more consequential. Moreover,
private DC pensions are becoming increasingly common and are predicted to grow
as current cohorts age \citep{cribb_karjalainen_ifs_2023}. Therefore, policies
regarding how private pensions can be accessed will have a larger impact on
overall welfare for retirees.

Moreover, understanding how retirees spend their money over retirement is an
important policy question of its own. If retirees spend too much in early
retirement, the state may need to provide for them towards the end of their
lives. This has implications for government budgets, especially since population
aging means more individuals may require expensive end of life care. On the
other hand, if retirees over-save and do not spend, perhaps in order to bequest
wealth, the economy may be dynamically inefficient -- a state in which the
capital stock is too high. In this case consumption, per capita could be
increased if savings were decreased.

\subsection{Literature review}
My paper draws on three main strands of literature: that of the annuity puzzle,
exploring reasons why retirees choose not to annuitise; the retirement
saving puzzle, which seeks to explain why retirees drawdown assets slowly; and
the lifecycle model, the predictions of which generate the above puzzles.

\cite{yaari_65} showed that under standard assumptions with risk averse agents
we would expect individuals to annuitise all of their wealth at retirement, so
that they are insured against the risk of long life. More recently,
\cite{davidoff_brown_diamond_aer_2005} show that complete annuitisation is
optimal under a broader range of assumptions. Since then, there has been much
discussion of possible reasons for why people choose not to annuitise.
\cite{finkelstein_porteba_2002, finkelstein_porteba_2004} find evidence of
adverse selection into the UK annuity market and \cite{mitchell_et_al_1999} find
evidence of the same in the US. This means longer-lived individuals select into
buying annuities, driving up annuity prices since providers need to pay
annuitants for longer than they originally predicted. This
makes the `money's worth' of an annuity lower for the general population as
opposed to the population of annuitants, putting off individuals from buying an
annuity. However, \cite{mitchell_et_al_1999} also find that theory would still
predict annuitisation because the money's worth of annuities is not that far
from `actuarially fair' (meaning the present discounted value of annuities
equals the price that an individual pays for one).

However, \cite{friedman_warshawsky_qje_1990} show that annuitisation decisions
can be fully explained by social security and `actuarially unfair' annuities, at
least for decisions in early retirement. They solve an augmented lifecycle model
and calculate annuity demand for a range of money's worths. For plausible values,
they find that individuals would optimally not annuitise much wealth. Similarly,
to \cite{finkelstein_porteba_2004}, \cite{friedman_warshawsky_chicago_1988} show
that there is a significant difference between the life expectancy of annuitants
and the general population in the American annuity market, which therefore
implies a degree of adverse selection. But, they find this cannot fully explain
the annuitisation puzzle, and only when bequest motives are added to the model
can annuitisation rates be rationalised.

More recently, \cite{lockwood_red_2012} builds on this by showing that a
realistic bequest motive in lifecycle simulations achieves realistic
annuitisation rates. He solves a simple lifecycle model with bequest motives
taken from several recent papers in the retirement savings literature. The
bequest motives he picks match other important aspects of the empirical
distribution of bequests, such as how much individuals bequeath and the
financial characteristics of these individuals. \cite{lockwood_aer_2018} also
finds that a model that treats bequests as luxury goods fits the lack of demand
for long-term care insurance well. \cite{odea_sturrock_rest_2023} find that
subjective life expectancies can explain a large part of the annuitisation
puzzle, especially if agents are less risk averse.

Apart from bequests and annuity prices, a range of other reasons for the annuity
puzzle have been explored. Precautionary savings for unexpected end-of-life
medical expenses might cause individuals to prefer holding assets rather than
an annuity. Moreover, \cite{vidalmelia_lejarragagarcia_munich_2004} suggest
that couples might share the risk of long life by pooling their resources, and
therefore couples may be less likely to annuitise their wealth. Another
explanation is that individuals already have enough guaranteed income from DB
pension schemes and state pensions that they do not need to annuitise more.

Given that uncertain end-of-life medical expenses are not such an issue in the
UK, as opposed to the US, and that the focus of the recent literature has been
on the bequest model and subjective life expectancy, I choose also to focus on
these explanations in the UK context. To my knowledge, there has been no attempt
to analyse the consumption response to the 2014 reform, nor evaluate predictions
from common lifecycle models for this particular policy change.

\section{Data and policy reform}

@@

@@

Successive governments in the UK have reformed both the public and
private pension system. Prior to 1987, participation in private schemes was
limited to employees of firms that had offered them, and there were few
alternatives to the public state pension or DB scheme that a public sector
employer would offer. From 1989, individuals in the UK were allowed to open tax
advantaged self-invested personal pensions alongside any pension their employer
offered. The 2004 Finance Act rationalised taxation rates on DB and DC pensions.
The rules for DC pension pots were such that an individual had to buy an annuity
after optionally taking a maximum of 25\% of the pot as a tax-free lump sum
withdrawal. DC pension pots were accessible from age 55, and most required that
they be accessed before the age of 75.

@@

The impact of the reforms on annuity demand has been documented by
\cite{cannon_et_al_nier_2016}. Figure \ref{fig:annovertime} shows
purchases of annuities over time and demonstrates the sharp decrease in purchases
that ocurred from 2014 to 2015. There was an increase in the number of pensions
that were being accessed using an income drawdown product, but these only partly
account for the drop in annuitisation as is shown by the graph.

    \centering
    \includegraphics[width=0.9\columnwidth]{figures/annuity_overtime.pdf}

    \caption{Product Sales Data}
    \label{fig:annovertime}
\end{figure}

@@

    \
\end{figure}
Figure \ref{fig:ann2122} shows the distribution of methods used to access pension pots
at retirement in 2021-22. In 2021, 196,736 pots were fully withdrawn at
retirement, accounting for over 50\% of pots. Prior to the policy reform this
was not the case -- most DC pension pots were accessed through
annuitisation.

Also of specific interest to the annuity market was the European Union's `Equal
Treatment in Goods and Services Directive' of 2004. This prohibited
discrimination in the provision and cost of goods and services based on sex. Up
until 2011, there was a clause that stated insurers were allowed to charge
different premiums if this was based on evidence that sex is correlated with
different amounts of risk. However, in March 2011, the European Court of Justice
ruled that insurers were not allowed to charge different amounts and gave them
until December 2012 to implement the ruling. This change meant that annuity
products could no longer be priced differently for men and women in the UK --
changing the money's worth of an annuity. I run a robustness test in the
empirical section to check whether this reform biases my results.

@@

\subsection{Covariate balance}

Table \ref{tab:sum_stats} shows various summary statistics for each variable of
interest from ELSA and the ONS.

An individual is defined as treated if they retired between and including 2015
and 2017, and an individual is in the control group if they retired in 2014 or
earlier. Since I am interested in consumption in early retirement, an
individuals interview date needs to be within the first three years of them
retiring.

The treated retirement group is smaller with just 303 non missing observations
for gender as opposed to 765 individuals in the control group. The control group
are more male, retired at a slightly younger age and expected to retire slightly
earlier. This could be because this period was affected by the increase in the
state pension age for women from age 60 in 2010 to age 65 in 2018, so early
cohorts had slightly earlier state pension eligibility ages. But, this change
happened gradually and ocurred over the whole period, so I do not expect it to
influence the results. The DifferenceAge row tracks the difference between
expected and actual retirement age.

The treatment group has higher financial wealth with a median of £67,000 at the
time of interview as opposed to £54,300 in the control group. Likewise, the
second group are more likely to have held a DC pension at some point and have
more money in them. Both groups are equally likely to have a DB pension with
roughly 45\% of individuals across both samples having a DB pension. In general,
DB pensions are more prevalent than DC pensions in the data -- this reflects the
trend that is observed in the general UK population. Home ownership is roughly
equal across groups although the second group has slightly higher housing
wealth.

@@

Table \ref{tab:cov_balance} shows these results. As expected, retirement year
and interview year are greater for the treated group. The difference between
expected and real retirement age is 0, which shows that individuals are not
delaying retirement, beyond their original expectation, in order to be part of
the treatment. DC pension value is higher, as is financial wealth but these are
both imprecisely estimated. Because of the differences in these variables I add
them to the regression model.

One threat to validity is manipulation of the running variable, retirement year.
As observed above in Tables \ref{tab:sum_stats} and \ref{tab:cov_balance}, it
does not appear as though individuals are delaying retirement. However, it is
possible that individuals do not buy an annuity at the time of retirement since
the law prior to 2014 only required that they access the pot by age 75. They may
decide to keep their DC pension untouched and live off other income and assets
before accessing it later. Since I do not have data for when individuals
purchase annuities, I cannot track whether annuity purchases were delayed for
people in the treatment group. However, there is relatively stable annuity
demand before the policy
reform. If individuals delayed annuity purchases in expectation of the reform,
so that they would not have to buy an annuity at all, we would observe declines
in purchases before the reform was announced. Likewise, if individuals bought an
annuity early in expectation that the end of the compulsory market would cause
prices to go up, we would observe a spike in annuity purchases prior to 2014.

Moreover, if individuals retired and then delayed purchasing an annuity, the
control group would have lower rates of annuitisation as well. This would bias
my results downward, since then individuals who retired in prior to the policy
change would also be consuming more.

@@

In this section, I outline the key empirical models I run, first with real
data from ELSA, and then with simulated data from the lifecycle models. I then
see which lifecycle model better fits the consumption response that occurred as
a result of the pension reform.

I use a fuzzy regression discontinuity design for which I compare the
consumption of individuals who retired after the policy reform to that of people
who retired just before the reform. The key assumption implicit in regression
discontinuities is that nothing else changes at the time of the jump apart from
the policy of interest, and that the policy occurs without individuals
predicting it and altering their behaviour. As I have argued above, the
demographic information in the data suggests that individuals did not delay
retirement and that there was no delay in annuitisation as shown by the quick
and sudden decline in annuity purchases. Moreover, anecdotal evidence from media
and business sources at the time of the announcement show that there was
surprise the government put forward the reform, with Money Marketing describing
the change as a ``bombshell" \citep{money_marketing_announcement}.

I use consumption data of individuals up to 3 years into their retirement so
that the sample size is larger. So, if someone retired in 2015 and had
consumption data in 2015 and 2017, I include both values.

@@

Table \ref{tab:DcOnlyRes} shows the results of this specification on the various
consumption indicators. Column 1 has total monthly consumption on the left hand
side of the regression equation. We can see that being in the treatment group is
associated with £36 more a week in spending overall. This is statistically
significant using robust standard errors clustered at the treatment level. Being
further into retirement at the time of interview is associated with higher total
consumption -- this is surprising since lifecycle models predict consumption
decreases over retirement and this has also been documented empirically
\citep{hurd_rohwedder_nber_2003}.

Columns 2 through 4 use food consumption as the outcome variable. Interestingly,
food consumption inside the house does not increase as a result of the policy
reform but food consumed away from the house does increase. A shock to income,
caused by no longer needing to annuitise, may cause spending on luxury goods to
increase and for spending on their substitutes, like food in the house relative
to food away from home, to decrease. This wealth effect may be driving the
results we observe in Table \ref{tab:DcOnlyRes} since food consumption outside the
home increases as does expenditure on clothing -- seen in column 5.

@@

The key identifying assumption for DD is the parallel trends assumption. In this
context, it means that consumption differences between the DB, DC and no pension
group across cohorts would have been the same were it not for the policy change.
Therefore for DD to be invalid, we would need a reason to expect that the
individuals with a DB and no pension who are retiring post 2014 are impacted by
some factor that did not equally impact the DC group. Given that there were no
other pension reforms at this time that differently affected these groups, I
claim that the parallel trends assumption holds.

Table \ref{tab:ElsaAllData} shows these results. The interactions at the bottom
of the table show us what happened to individuals with a given pension type
relative to the reference cohort, which has no pension. Those with only a DC
pension had an increase in total consumption which was higher than those with a
DB pension and higher than those with no pension. Consumption on food decreased
slightly for the DC group relative to no pension, but the decrease was less than
that for the DB group. Expenditure on clothing increased for the DC group
relative to the no pension and DB pension groups. The increase in sample size
also brings decreases in the standard errors of these variables -- the
interactions are now all significant a the 5\% level as is the main coefficient
on retiring post-reform.

However, it is worth noting that there is a significant change between the DB
and the no pension group. A natural test of the parallel trends assumption is to
see whether two different, untreated, units have different outcomes
post-treatment. If they are different post treatment then there could be
something else driving the differences between the treatment group and the
non-treated groups. This therefore raises questions about the validity of the DD
regression in this setting.

We can also test whether treatment intensity is correlated with larger increases
in spending. Those who have more money in a DC pension would be impacted by the
policy to a greater extent than those with smaller pensions, since they would be
forced to annuitise a larger amount. So, I interact the policy with DC wealth.
As in Table \ref{tab:DcOnlyRes}, this regression only uses those individuals who
have a DC pension.

@@

Table \ref{tab:DcOnlyInteract} shows the results. The interaction with DC
pension size is positive, meaning that a larger DC pension pot is associated with a
greater increase in total monthly consumption. For the all other categories,
more pension wealth is associated with lower consumption, though for food out
the house and for clothing the main coefficients are still positive and
significant.

As mentioned above, the DC pension pot value data after
wave 5 has not been released yet, so for individuals who retire after 2011 I
have no DC wealth variable. For Table \ref{tab:DcOnlyInteract}, pension wealth
from the last available year for an individual was given a real return of 3\%
and compounded until year of interview. Given that this is an imperfect measure,
I also run the regression with financial wealth interacted with the treatment
variable.

Table \ref{tab:DcOnlyFinWealthInteract} contains these results and shows a
similar pattern to Table \ref{tab:DcOnlyInteract}. The reform is associated with
an increase in total monthly consumption and more financial wealth increases the
size of this effect. As for the DC pension value variable, the interactions
in the regressions of the other consumption variables have negative signs.

\begin{landscape}
    \linespread{1}
    \input{tables/elsa_results_only_dc.tex}
\end{landscape}

\begin{landscape}
    \linespread{1}
    \input{tables/elsa_results_all_with_interact.tex}
\end{landscape}
\begin{landscape}
    \linespread{1}
    \input{tables/elsa_results_dc_pot_interact.tex}
\end{landscape}

\begin{landscape}
    \linespread{1}
    \input{tables/elsa_results_fin_wealth_interact.tex}
\end{landscape}
\begin{landscape}
    \linespread{1}
    \input{tables/rob_elsa_results_only_dc_only_exp.tex}
\end{landscape}

@@

\subsection{Robustness checks}

I carry out two checks for robustness. Firstly, I exclude those individuals who
retired more than one year away from their expected retirement year -- I define
an individual's expected retirement year as their expected retirement age in
their first interview with ELSA, added to the year that the interview was
carried out in. This checks whether the effects shown above are being driven by
individuals either retiring early in order to buy an annuity, since one may be
cheaper pre-reform because the compulsory market reduces adverse selection
\cite{finkelstein_porteba_2002}, or retiring later to avoid annuitisation
altogether.

I also check whether the EU reform to annuity pricing is changing the results in
some way. Since the reform only came into effect in December 2012, there are some
retirees in the sample who may have purchased an annuity before
annuity pricing was made the same for men and women. I drop anyone with a
retirement year in 2012 or 2011 and repeat the analysis from Table
\ref{tab:DcOnlyRes}.
